***
# Introduction
**Scripts used to process and analyze hMeDIP and MeDIP-seq data**

## Packages
```{bash}
# Biowulf packages 
trimgalore/0.6.6
fastqc/0.11.9
multiqc/1.9
bowtie/2
samtools/1.9
picard/2.23.7
bedtools/2.30.0
deeptools/3.5.0
```

```{r}
# R packages
library(tidyverse)
library(viridisLite)
library(viridis)
library(ggpubr)
library(corrplot)
```

# Data Pre-processing
```{bash}
##== linux command ==##
path="/path"

#text file with the names of all the samples
#hMeDIP_Names.txt
```

## Adapter trimming
```{bash}
##== linux command ==##
mkdir -p ${path}/fastq/adapt_trim

for Sample_names in $(cat Sample_names.txt)
do
  echo "fastq: " $Sample_names
  trim_galore ${path}/fastq/${Sample_names}_R?.fastq.gz --illumina -o ${path}/fastq/adapt_trim/
done
```

## FastQC on resulting adapter trimmed FastQ files
```{bash}
##== linux command ==##
mkdir -p ${path}/fastq/adapt_trim/fastqc
TMP=/lscratch/$SLURM_JOB_ID

for Sample_names in $(cat Sample_names.txt)
do
  fastqc -o ${path}/fastq/adapt_trim/fastqc/ -d $TMP -t 12 --noextract \
  -f fastq ${path}/fastq/adapt_trim/${Sample_names}_R?.fq.gz
done
```

## MultiQC to summarize QC data
```{bash}
##== linux command ==##
multiqc ${path}/fastq/adapt_trim/fastqc/ -n ${path}/fastq/adapt_trim/fastqc/multiqc_hMeDIP
```

# Genome Alignment
## Bowtie2 alignment to mm10
```{bash}
##== linux command ==##
mkdir -p ${path}/bam
mkdir -p ${path}/bam/bowtie2_summary
mkdir -p ${path}/bam/fragmentLen

#mm10 bowtie2 index
export BOWTIE2_INDEXES=/fdb/igenomes/Mus_musculus/UCSC/mm10/Sequence/Bowtie2Index/

for Sample_names in $(cat Sample_names.txt)
do
  bowtie2 -p 24 --end-to-end --no-mixed --no-discordant --very-sensitive -I 10 -X 700 \
  --phred33 -x genome -1 ${path}/fastq/adapt_trim/${Sample_names}_R1.fq.gz -2 ${path}/fastq/adapt_trim/${Sample_names}_R2.fq.gz \
  -S ${path}/bam/${Sample_names}.sam &> ${path}/bam/bowtie2_summary/${Sample_names}_bowtie2.txt 
  
  #Extract the 9th column from the alignment sam file which is the fragment length
  samtools view -@ 24 -h -F 4 -q 10 -bS ${path}/bam/${Sample_names}.sam | \
  awk -F'\t' 'function abs(x){return ((x < 0.0) ? -x : x)} {print abs($9)}' | \ 
  sort | uniq -c | awk -v OFS="\t" '{print $2, $1/2}' > ${path}/bam/${Sample_names}_fragmentLen.txt
 
  #convert sam to filtered bam
  samtools view -@ 24 -h -F 4 -q 10 -bS ${path}/bam/${Sample_names}.sam > ${path}/bam/fragmentLen/${Sample_names}_filtered.bam
  rm ${path}/bam/${Sample_names}.sam
done
```

## Report and Visualize mapping summary
### Sequencing depth
```{r}
##=== R command ===## 
ageList = c("Y", "O")
sampleList = c("5hmC", "input")
repList = c("1","2","3", "4")
projPath = "/path"
```

### Bowtie2 results summary
```{r}
##=== R command ===## 
alignResult = c()
for(age in ageList){
  for(sample in sampleList){
    for(rep in repList){
  alignRes = read.table(paste0(projPath, "/bowtie2_summary/", age, "_", sample, "_", rep, "_bowtie2.txt"), header = FALSE, fill = TRUE)
  alignRate = substr(alignRes$V1[6], 1, nchar(as.character(alignRes$V1[6]))-1)
  alignResult = data.frame(Sample = sample, Replicate = rep, Age = age,
                           SequencingDepth = alignRes$V1[1] %>% as.character %>% as.numeric, 
                           MappedFragNum_mm10 = alignRes$V1[4] %>% as.character %>% as.numeric + alignRes$V1[5] %>% as.character %>% as.numeric, 
                           AlignmentRate_mm10 = alignRate %>% as.numeric)  %>% rbind(alignResult, .)
    }}}
alignResult$Sample = factor(alignResult$Sample, levels = sampleList)
alignResult %>% mutate(AlignmentRate_mm10 = paste0(AlignmentRate_mm10, "%"))
```

### Sequencing depth boxplot
```{r}
##=== R command ===## 
#Figure theme
theme_USGS_box <- function(base_family = "sans", ...){
  theme_bw(base_family = base_family, ...) +
  theme(
    panel.grid = element_blank(),
    legend.background = element_rect(color = "black"),
    axis.text.y = element_text(margin=unit(c(0.3,0.3,0.3,0.3), "cm")), 
    axis.text.x = element_text(margin=unit(c(0.05,0.05,0.05,0.05), "cm"))
  )
}

#Young as base level
alignResult$Age <- as.factor(alignResult$Age)
alignResult$Age <- relevel(alignResult$Age, ref = "Y")

pd = position_dodge(width = 0.85)

#Figures
fig1_hMe <- alignResult %>% ggplot(aes(x = Sample, y = SequencingDepth/1000000, fill =Age)) +
  geom_boxplot(position=pd, coef=6) +
  expand_limits(y = 0) +
  geom_point(pch = 21, position = position_jitterdodge())+
  scale_fill_manual(values= c("darkgray", "red"), 
                       name="Age",
                       breaks=c("Y", "O"),
                       labels=c("Young", "Old")) +
  theme_USGS_box(base_size = 15) + 
  ylab("Sequencing Depth per Million") +
  xlab("") + 
  ggtitle("A. Sequencing Depth")

fig2_hMe <- alignResult %>% ggplot(aes(x = Sample, y = MappedFragNum_mm10/1000000, fill = Age)) +
  geom_boxplot(position=pd, coef=6) +
  expand_limits(y = 0) +
  geom_point(pch = 21, position = position_jitterdodge())+
  scale_fill_manual(values= c("darkgray", "red"), 
                       name="Age",
                       breaks=c("Y", "O"),
                       labels=c("Young", "Old")) +
  theme_USGS_box(base_size = 15) +
  ylab("Mapped Fragments per Million") +
  xlab("") +
  ggtitle("B. Alignable Fragment (mm10)")

fig3_hMe <- alignResult %>% ggplot(aes(x = Sample, y = AlignmentRate_mm10, fill = Age)) +
  geom_boxplot(position=pd, coef=6) +
  expand_limits(y = 0) +
  geom_point(pch = 21, position = position_jitterdodge())+
  scale_fill_manual(values= c("darkgray", "red"), 
                       name="Age",
                       breaks=c("Y", "O"),
                       labels=c("Young", "Old")) +
  theme_USGS_box(base_size = 15) +
  ylab("% of Mapped Fragments") +
  xlab("") +
  ggtitle("C. Alignment Rate (mm10)")
```

```{r}
#Statistical test
hmC <- subset(alignResult, Sample == "5hmC")
Input <- subset(alignResult, Sample == "input")

#Seq depth
t.test(hmC$SequencingDepth ~ hmC$Age)
t.test(Input$SequencingDepth ~ Input$Age)

#Mapped Fragment number
t.test(hmC$MappedFragNum_mm10 ~ hmC$Age)
t.test(Input$MappedFragNum_mm10 ~ Input$Age)

#Alignment rate
t.test(hmC$AlignmentRate_mm10 ~ hmC$Age)
t.test(Input$AlignmentRate_mm10 ~ Input$Age)
```

### Report and Visualize fragment length summary
```{r}
##=== R command ===## 
## Collect the fragment size information
fragLen = c()
for(age in ageList){
  for(sample in sampleList){
    for(rep in repList){
        fragLen = read.table(paste0(projPath, "/fragmentLen/", age, "_", sample, "_", rep, "_fragmentLen.txt"), header = FALSE) %>% 
        mutate(fragLen = V1 %>% as.numeric, fragCount = V2 %>% as.numeric, Weight = as.numeric(V2)/sum(as.numeric(V2)), Sample = sample, Replicate = rep, Age = age) %>% 
        rbind(fragLen, .) 
    }}}

#Young as base level
fragLen$Age <- as.factor(fragLen$Age)
fragLen$Age <- relevel(fragLen$Age, ref = "Y")

fig5A = fragLen %>% ggplot(aes(x = Sample, y = fragLen, weight = Weight, fill = Age)) +
    geom_violin(bw = 5) +
    scale_y_continuous(breaks = seq(0, 800, 50)) +
    scale_fill_manual(values= c("darkgray", "red"), 
                       name="Age",
                       breaks=c("Y", "O"),
                       labels=c("Young", "Old")) +
    theme_USGS_box(base_size = 15) +
    ggpubr::rotate_x_text(angle = 20) +
    ylab("Fragment Length") +
    xlab("")
```

```{r}
#Subset to have hmC and input in different groups
hmC_frag <- subset(fragLen, Sample == "5hmC")
Input_frag <- subset(fragLen, Sample == "input")

#Alignment rate
t.test(hmC_frag$fragLen ~ hmC_frag$Age)
t.test(Input_frag$fragLen ~ Input_frag$Age)
```

#Sorting, keeping only uniquely mapped reads, and removing duplicates
```{bash}
##== linux command ==##

for Sample_names in $(cat Sample_names.txt)
do 
  sambamba sort -t 2 -o ${path}/bam/${Sample_names}_sorted.bam ${path}/bam/${Sample_names}_filtered.bam
  sambamba view -h -t 2 -f bam -F "[XS] == null and not unmapped and not duplicate" \
  ${path}/bam/${Sample_names}_sorted.bam > ${path}/bam/${Sample_names}_unique_noDUP.bam
  
  ##To remove filtered and sorted bam files
  rm ${path}/bam/${Sample_names}_filtered.bam
  rm ${path}/bam/${Sample_names}_sorted.bam
done
```

### Report and visualize the duplication rate and unique library size 
#### Duplication results summary
```{r}
##=== R command ===## 
dupResult = c()
for(age in ageList){
  for(sample in sampleList){
  for(rep in repList){
  dupRes = read.table(paste0(projPath, "/picard_summary/", age, "_", sample, "_", rep, "_noDUP.txt"), header = #TRUE, fill = TRUE)
  dupResult = data.frame(Sample = sample, Replicate = rep, Age=age, MappedFragNum_mm10 = dupRes$READ_PAIRS_EXAMINED[1] %>% as.character %>% 
  as.numeric, DuplicationRate = dupRes$PERCENT_DUPLICATION[1] %>% as.character %>% as.numeric * 100, EstimatedLibrarySize = dupRes$ESTIMATED_LIBRARY_SIZE[1] %>% as.character %>% as.numeric) %>% 
  mutate(UniqueFragNum = MappedFragNum_mm10 * (1-DuplicationRate/100))  %>% rbind(dupResult, .)
}}}
dupResult
```

#### Duplication boxplots
```{r}
##=== R command ===## 
#Young as first
dupResult$Age <- as.factor(dupResult$Age)
dupResult$Age <- relevel(dupResult$Age, ref = "Y")

pd = position_dodge(width = 0.85)

fig4_hMe = dupResult %>% ggplot(aes(x = Sample, y = DuplicationRate, fill = Age)) +
  geom_boxplot(position=pd) +
  expand_limits(y = 0) +
  geom_point(pch = 21, position = position_jitterdodge())+
  scale_fill_manual(values= c("darkgray", "red"), 
                       name="Age",
                       breaks=c("Y", "O"),
                       labels=c("Young", "Old")) +
  theme_USGS_box(base_size = 15) +
  ylab("Duplication Rate (%)") +
  xlab("") 

fig5_hMe= dupResult %>% ggplot(aes(x = Sample, y = EstimatedLibrarySize, fill = Age)) +
  geom_boxplot(position=pd) +
  expand_limits(y = 0) +
  geom_point(pch = 21, position = position_jitterdodge())+
  scale_fill_manual(values= c("darkgray", "red"), 
                       name="Age",
                       breaks=c("Y", "O"),
                       labels=c("Young", "Old")) +
  theme_USGS_box(base_size = 15) +
  ylab("Estimated Library Size") +
  xlab("") 


fig6_hMe = dupResult %>% ggplot(aes(x = Sample, y = UniqueFragNum, fill = Age)) +
  geom_boxplot(position=pd, coef=6) +
  expand_limits(y = 0) +
  geom_point(pch = 21, position = position_jitterdodge())+
  scale_fill_manual(values= c("darkgray", "red"), 
                       name="Age",
                       breaks=c("Y", "O"),
                       labels=c("Young", "Old")) +
  theme_USGS_box(base_size = 15) +
  ylab("# of Unique Fragments") +
  xlab("")
```

## Remove ENCODE blacklisted regions from Bam files
```{bash}
wget -O ${path}/bam/ENCFF547MET.bed.gz https://www.encodeproject.org/files/ENCFF547MET/@@download/ENCFF547MET.bed.gz
gunzip ${path}/bam/ENCFF547MET.bed.gz
BLACKR=${path}/bam/ENCFF547MET.bed

for Sample_names in $(cat Sample_names.txt)
do
  bedtools intersect -a ${path}/bam/${Sample_names}_noDUP.bam -b $BLACKR -v > ${path}/bam/${Sample_names}_noDUP_B.bam
  rm ${path}/bam/${Sample_names}_noDUP.bam 
done
```

# Generate browser tracks
## Index bam files
```{bash}
for Sample_names in $(cat Sample_names.txt)
do
  samtools index ${path}/bam/${Sample_names}_noDUP_B.bam
done
```

## Generate bigWig files
```{bash}
mkdir -p ${path}/bigwig

for Sample_names in $(cat Sample_names.txt)
do
  bamCoverage -b ${path}/bam/${Sample_names}_noDUP_B.bam -o ${path}/bigwig/${Sample_names}.bw -of bigwig \
  --effectiveGenomeSize 1870000000 --extendReads 200 --normalizeUsing RPKM -p $SLURM_CPUS_PER_TASK
done
```

# Assess replicate reproducibility 
```{bash}
INPUT1=${path}/bigwig/*.bw

#Creating npz file
multiBigwigSummary bins -b $INPUT1 -o ${path}/bigwig/hMeDIP_bigwig_summary.npz -p max

#PCA plot
plotPCA -in ${path}/bigwig/hMeDIP_bigwig_summary.npz -o hMeDIP_PCAsummary.pdf
```


